---
title: "Practicals 03: Rmarkdown and testing"
subtitle: "Reproducible science with R"
author: "January Weiner"
date: "`r Sys.Date()`"
output:
    html_document:
      toc: true
      toc_float: true
outputoff:
  xaringan::moon_reader:
    self-contained: true
    css: ["default", "files/cubi-fonts.css", "files/style.css" ]
    lib_dir: libs
    nature:
      highlightStyle: github
      countIncrementalSlides: false
      ratio: "16:9"
      beforeInit: "files/macros.js"
toc: no
---

```{r,echo=FALSE}
## Set default options for the knitr RMD processing
knitr::opts_chunk$set(echo=FALSE,warning=FALSE,message=FALSE,fig.width=5,fig.height=5,cache=TRUE,autodep=TRUE, results="hide")
```

```{r libraries,cache=FALSE,echo=FALSE,warning=FALSE,message=FALSE,error=FALSE}
library(tidyverse)
library(cowplot)
```


# Exercises 1.

 1. Using the WIID data set, reproduce the example from lecture for another
    selection of years / regions (maybe not for countries, but for regions
    or subregions?) Can you think of another representation?
 2. Using the mtcars data set and any graphics system, create a box plot 
    showing mpg depending on the number of cylinders (cyl). The latter
    should be treated as a category, not a continuous variable! (in ggplot:
    use the `geom_boxplot()` geom).

# Exercises 2.

 3. Load the [transcriptomics_results.csv](https://raw.githubusercontent.com/january3/ReproducibleR2021/master/Datasets/transcriptomics_results.csv) data set. 
    This is a data set derived from
    [this](https://www.nature.com/articles/s41598-019-56994-8) paper,
    showing transcriptomic responses to the vaccination at different
    timepoints. Using whatever tools (base R, `ggplot() + geom_point()`)
    plot the log fold changes on day 1 vs log fold changes on day 2. 
 4. Create an additional column to the transcriptomic data set. Fill it with
    either "TRUE" (a gene is significant on day 1) or "FALSE" (a gene is not
    significant). Define significance in terms of log fold change *and*
    p-value (i.e., a gene is significant if the absolute lfc is above a
    threshold *and* p-value is below a threshold).
 5. Create a volcano plot (i.e., log fold change on x axis, -log10(p-value)
    on y axis) showing in red genes which are significant on day 1. (for
    example, with ggplot2, use the color esthetics with the "sign" column
    you created in the previous step)
 6. Using a similar approach, color (or symbol) code genes which are:
     * not significant on Day 1 *and* Day 2
     * significant only on Day 1
     * significant only on Day 2
     * significant on both days

