---
title: "Practicals 01: Working with R"
subtitle: "Reproducible science with R"
author: "January Weiner"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    self-contained: true
    css: ["default", "files/cubi-fonts.css", "files/style.css" ]
    lib_dir: libs
    nature:
      highlightStyle: github
      countIncrementalSlides: false
      ratio: "16:9"
      beforeInit: "files/macros.js"
toc: no
---

```{r,echo=FALSE}
## Set default options for the knitr RMD processing
knitr::opts_chunk$set(echo=FALSE,warning=FALSE,message=FALSE,fig.width=5,fig.height=5,cache=TRUE,autodep=TRUE, results="hide")
```

```{r libraries,cache=FALSE,echo=FALSE,warning=FALSE,message=FALSE,error=FALSE}
library(tidyverse)
library(cowplot)
```

## Repetitorium of day 1: Example

First, complete a short test:
[https://forms.gle/btAkm8qYfA6fuF3i8](https://forms.gle/btAkm8qYfA6fuF3i8)

---

## Repetitorium of day 1: Example

There is a lake in a garden. Every day, the water lilies cover twice as
much area as the previous day. On the first day, the water lilies cover
1/100th of the area of the lake.

 * What is the formula to calculate the area covered by the water lilies on
   day n? (pen and paper, not in Rstudio!)
 * assign days 1 ... 10 to variable x. Now assign the fraction of the area
   covered by water lilies on day n to variable y
 * plot x vs y (use the simple `plot()` function)
 * When will the half of the area be covered by water lilies? Use `abline(h=.5)`
   (what does it do?) to show a graphical solution
 * What is the fraction on day 3? Hint: what does `y[3]` do?
 * Speaking of which, what dows `y[4:5]` do? (just try it!)

---

## Exercise

 * Make the plot yourself.
 * use 'col' parameter of `plot` and a color name (e.g. "red") to change
   the color of the line (`plot(..., col="red")`)
 * what if each day the lilies cover a fraction of area that is 1.5 the
   fraction of area on the previous day?
 * Use `lines(x, y)` function to put a second line on the plot

---

## Creating your own functions for fun and profit

 * If you have to write something more than twice, you should turn it into
   a function!
 * R is a *functional* language

---

## Creating a function

```{r eval=FALSE,echo=TRUE}
#' Function name
#' Function description
some_name <- function(param1, param2=2) {

## code comment
# <your code goes in here>

}
```

---

## Example

```{r eval=FALSE,echo=TRUE}

#' Function for pond calculation
pond_surface_covered <- function(init_pond_surface, day=0) {

  ## "**" and "^" do the same thing
  ret <- init_pond_surface * 2^day

  return(ret)
}
```

---

## Example


```{r eval=FALSE,echo=TRUE}

#' Generate a random sample from a given distribution
#' Usage: randsamp(n) where n is the size of the sample

randsamp <- function(n, d="uniform") {

  if(d == "uniform") {
    ret <- runif(n)
  } else if(d == "normal") {
    ret <- rnorm(n)
  } else {
    stop("I don't know this distribution!")
  }

  return(ret)
}
```

---

## Exercise 2.1

Take the function above and add a new option: the binomial distribution.
Which function can give you that? Search for a suitable function.

Note: binomial distribution is essentially what you get when you throw a
coin a number of times and add the results (1 for heads, 0 for tails). 
This function requires two more parameters: number of "coin tosses"
and probability of "heads". You need to adapt the code accordingly: choose
something meaningful, e.g. size=1 and prob=0.5 to simulate a single fair
coin toss.

---

# Reading and writing data

---

## Reading data

Main data types you will encounter:

---------------------------- ------------------------------- --------------------------
Data type                    Function                        Notes
---------------------------- ------------------------------- --------------------------
TSV / TAB separated values   `read_table()`                  Tab is `\t`.
                                                             A very general and 
                                                             customizable function

CSV / comma separated        `read_csv()`                    a wrapper around
                                                             `read.table()`

XLS (old Excel)              `read_xls()`                    Just don't use it.
                             `read_excel()`                  From the
                                                             `readxl`
                                                             package.

XLSX (new Excel)             `read_xlsx()`                   From the `readxl` package.
                             `read_excel()`                  You need to
                                                             provide the
                                                             sheet number
                                                             you wish to
                                                             read. 
                                                             Note: returns a tibble,
                                                             not a data frame! 
---------------------------- ------------------------------- --------------------------

---

## Reading with GUI

You can use the "Files" pane to import some data types. When you do, you
get an import dialog to choose certain options; but then, the actual
command is copied to your console – you can copy and paste it and it will
work exactly the same.

---

## Tibbles / readxl

tibbles belong to the tidyverse. They are nice to work with and very
useful, but a bit different from regular data frames. To convert a tibble
to a data frame, you can do

```{r eval=FALSE}
mydataframe <- as.data.frame(read_xlsx("file.xlsx"))
```

One crucial difference between tibble and data frame is that `tibble[ , 1 ]` 
returns a tibble, while `dataframe[ , 1]` returns a vector. The second
crucial difference is that it does not support row names (on purpose!).

---

## Cleaning up data

`unique` returns the unique values in a vector. Use it to check the
correctness of the columns.

Use `class()` to make sure that the columns are what you think they are.

Use `summary()` to gain an overview of the data.

Convert with `factor` when necessary (and only then).

Also, you can use my function `colorDF_summary` from the `colorDF` package
(install from CRAN: `install.packages("colorDF")`)

---

## Excercise 2.2

Read, inspect and correct (if reasonable) the following files:

 * `cars.xlsx`
 * `iris.csv`
 * `expression_data_vaccination_example.xlsx` (large!)

The function `readxl_example("deaths.xls")` returns a file name. Read this
file. How can you omit the lines at the top and at the bottom of the file?
(hint: `?read_xls`). How can you force the date columns to be interpreted
as dates and not numbers?


